A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-08 analysis/paper/paper.Rmd
edit to reduce word counts

<!---network analysis and the basic concepts-->
Different from a static characterization of burial variables, network modeling approaches have the potential to study social changes in a dynamic way by examining the intercorrelations between burials. Drawing largely on network science and graph theory, a key assumption in the interpretation of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A network is viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Burials with the same kinds of prestige goods could indicate some underlying social relations where individuals share similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. Among the network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its applications [@Caimo2014]. In this paper, we try to go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections between burials and changes in network structures by examining evidence from a case study in northeastern Taiwan.

<!--research aim and questions with archaeological specific case study-->
We study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan based on the theoretical framework of corporate-network strategies [@Blanton1996; @Feinman2000]. We test our hypothesis that changes in power structure from a corporate mode to a network mode can be observed in the burial network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming the unequal social positions represented that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context. In addition, we introduce a novel Bayesian inference on statistical modeling for networks that is suitable for networks with smaller sample sizes.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups (represented by higher transitivity and negative popularity) toward a more centralized network with concentration of connections among fewer burials (lower transitivity and positive popularity) after the European arrival in the 17th century. This supports our models of chronological change from a more corporate-based to a more networked-based society, as indicated in the grave goods and burial attributes. These observations are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000; @Feinman2000political]. Compared to traditional models of vertical hierarchies that stress stratified social characteristics, the corporate-network model may be more relevant for understanding how a small scale society was organized, such as we see here in the pericolonial context of Kiwulan. A corporate-based society, such as the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Siegel1999]. In contrast, a network-based society, as suggested the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate/network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages.

<!--archaeological implication based on the results-->
The covariants for networks provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:table-posterior-stats)). In other words, we can know what kinds of socio-cultural factors associated with the acquisition and distribution of foreign prestige goods reflected by burial networks. Our covariants were age, sex, ritual pottery, and wealth ranks calculated based on the value of grave goods. The results indicate that the major factor contributing to ties between burials is ritual for the pre-European network, and wealth for the post-European network. Moreover, in the post-European network, burials with ritual vessels or same sex tend not to be tied that suggest they are not associated with the network pattern we observed. Those results illustrate that different social and economic mechanisms determined relations between burials, or the interconnections of prestige goods, for the two periods. Before European arrival, ritual behaviors seem to have been important factors that structure the interconnections of prestige goods in the burial context. This might imply that ritual could be a status indicator when the foreign goods were first introduced to the local society. After European presence was established, status indicators shifted from ritual to wealth differences with the least sex status discrimination. Moreover, geographical proximity, implying some kinship association, is not a factor for sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for the pre-European network and only sex shows a negative effect for the post-European network. This may also support a shift to more wealth and individual-based status competition without sex differences in the later phase. But we also note that no effects may be due to the unavailability of age and sex values for many burials in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on the assumption that burials can represent social structures [@Binford1971; @Saxe1970]. We argue that the social relation at Kiwulan changed because of the contact with European colonial power and associated foreign trade. This argument is supported by the changes in burial network structures from before and after European arrival and their associated underlying mechanisms. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation at Kiwulan in a pericolonial situation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, which supports a connection between a burial and the social role of the living person. Since the burial goods used for making ties were treated as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods observed in the burial data likely reflect social contexts of the living people at Kiwulan in a pericolonial context.

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, changes in interburial networks may result from differential access to exotic trade enabled by European contact, gradually leading to increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in local Indigenous culture where their values were amplified with concepts of wealth or power, since their values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals competing with each other for traditional goods and accumulating wealth. It also demonstrates that local Indigenous societies could actively incorporate and manipulate trade items in a pericolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. We recognize that the meaning of the burial goods could be refined with more ethnographic data to support our interpretations. Despite fragmentation in our archaeological data, such as missing age and sex attributes for some burials at Kiwulan, we can still capture the general patterns of network structures related to social implication [eg. @Erik2015]. Human skeletal remains can provide information about health for interpreting social status; however, the preservation of human remains at Kiwulan was generally not good to yield useful data. By focusing on burial goods, we seek to understand the social changes at Kiwulan in relation to the contact of Europeans.

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a novel approach for studying burials to interpret social structures using ERGMs within a Bayesian framework. We examined social changes in a European pericolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our model that the relationship between burials changed after the European colonization period in the 17th century. Before the arrival of Europeans, the burial network has a tendency of more clustered subgroups with pottery as ritual practice as the key formation mechanism. After European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual practice connections to wealth level connections suggest increased behavior of wealth accumulation stimulated by the European presence and associated long-distance trade network. This aligns with changes in horizontal hierarchy from a more corporate to a more network mode that demonstrates different degrees of social inequality in relation to different strategies for achieving power [@Feinman2000]. Using burial data with historical documents, we are able to detect changes in Indigenous societies indicative of increased social inequality to better understand indirect European colonial impacts that have been underestimated in this region [e.g. @Acabado2017; @Trabert2017]. Furthermore, our study also highlights the possibility of uneven access to foreign trade goods that allows active accumulation or even manipulation their values by local people instead of being passive recipients of trade goods.

>>>>

<!---network analysis and the basic concepts-->
Different from a static characterization of burial variables, network modeling approaches have the potential to study social changes in a dynamic way by examining the intercorrelations between burials. Drawing largely on network science and graph theory, a key assumption in the interpretation of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A network is viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Among the network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its applications [@Caimo2014]. In this paper, we try to go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections between burials and changes in network structures by examining evidence from a case study in northeastern Taiwan.

<!--research aim and questions with archaeological specific case study-->
We study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan based on the theoretical framework of corporate-network strategies [@Blanton1996; @Feinman2000]. We test our hypothesis that changes in power structure from a corporate mode to a network mode can be observed in the burial network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming the unequal social positions represented that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups toward a more centralized network with concentration of connections among fewer burials after the European arrival in the 17th century. This supports our pericolonial impact model that Kiwulan changed from a more corporate-based to a more networked-based society, as indicated in the grave goods and burial attributes. These observations are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000; @Feinman2000political]. Compared to vertical hierarchies that stress stratified social characteristics, the corporate-network strategies may be more relevant for understanding how a small scale society was organized, such as we see here in the pericolonial context of Kiwulan [cf. @Ueda2016]. A corporate-based society, such as the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Siegel1999]. In contrast, a network-based society, as suggested the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate-network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages.

<!--archaeological implication based on the results-->
The covariants for networks provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:table-posterior-stats)). In other words, we can know what kinds of socio-cultural factors associated with the acquisition and distribution of foreign prestige goods. Our covariants were age, sex, ritual pottery, and wealth ranks calculated based on the value of grave goods. The results indicate that the major factor contributing to burial networks is ritual for the pre-European network, and wealth for the post-European network. Moreover, in the post-European network, burials with ritual vessels or same sex tend not to be tied that suggest they are not associated with the network pattern we observed. Those results illustrate that different social and economic mechanisms determined relations between burials, or the interconnections of prestige goods, for the two periods. Before European arrival, ritual behaviors seem to have been important factors that structure the interconnections of prestige goods in the burial context. This might imply that ritual could be a status indicator when the foreign goods were first introduced to Indigenous societies. After European presence was established, status indicators shifted from ritual to wealth differences with the least sex status discrimination. Moreover, geographical proximity, implying some kinship association, is not a factor for sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for the pre-European network and only sex shows a negative effect for the post-European network. This may also support a shift to more wealth and individual-based status competition without sex differences in the later phase. But we also note that no effects could be due to the unavailability of age and sex values for many burials in our sample.

<!-- other evidence -->
We used burial data as a proxy to explore social relations reflected by interconnections of prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks according to their status [eg. @Coward2013, pp. 252]. We argue that the social relation at Kiwulan changed because of the contact with European colonial power and associated foreign trade. This argument is supported by the changes in burial network structures through time and their associated underlying mechanisms. It should be noted that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982]. However, this issue can be reduced by comparing the results with evidence from other archaeological contexts, such as residential areas [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution when the Europeans were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with the result of a more centralized burial network after the presence of Europeans, that supports a connection between a burial and the social role of the living person. Also, the burial goods used for making ties were treated as prestige goods or status items in the local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations represented by the flow of goods observed in the burial data likely reflect social contexts of the living people at Kiwulan.

<!--anthopological implication at Kiwulan and limitation-->
At Kiwulan, changes in interburial networks may result from differential access to exotic trade enabled by European contact, gradually leading to increased social differentiation. The image of European power could be embedded into prestige objects pre-existing in local Indigenous culture where their values were amplified with concepts of wealth or power, since their values are contextually mutable and entangled in historical processes [@Aswani2003; @Thomas2009]. This may lead to ambitious individuals competing with each other for traditional goods and accumulating wealth. It also demonstrates that local Indigenous societies could actively incorporate and manipulate trade items in a pericolonial context with weak colonial control, gradually resulting in social changes from a more corporate to a more network strategy. We recognize that the meaning of the burial goods could be refined with more ethnographic data to support our interpretations. Human skeletal remains can provide information about health for interpreting social status; however, the preservation of human remains at Kiwulan was generally not good to yield useful data. By focusing on burial goods, we seek to understand the social changes at Kiwulan in relation to the contact of Europeans. Despite fragmentation in our archaeological data, such as missing age and sex attributes for some burials at Kiwulan, we can still capture the general patterns of network structures related to social implication.

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a novel approach for studying burials to interpret social structures using ERGMs within a Bayesian framework. We examined social changes in a European pericolonial context in northeastern Taiwan. We tested the hypotheses of changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our model that the relationship between burials changed after the European colonization period in the 17th century. Before the arrival of Europeans, the burial network has a tendency of more clustered subgroups with pottery as ritual practice as the key formation mechanism. After European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual practice connections to wealth level connections could suggest increased behavior of wealth accumulation stimulated by the European presence and associated long-distance trade network. This aligns with changes in horizontal hierarchy represented by corporate-network societies, that demonstrates different degrees of social inequality in relation to different strategies for achieving power [@Feinman2000]. Using burial data with historical documents, we are able to detect changes in Indigenous societies indicative of increased social inequality to better understand indirect European colonial impacts that have been underestimated in this region [e.g. @Acabado2017; @Trabert2017].
